Using SMT engine to generate Symbolic Automata\thanks{This work was partially funded by the Associated Team FM4CPS between INRIA and ECNU, Shanghai}

### 2018-01-12 2017-Xudong-SMT/FM-2018-extended.tex
eric

\title{Using SMT engine to generate Symbolic Automata\thanks{This work was partially 
funded by the Associated Team FM4CPS
  between INRIA and ECNU, Shanghai}}
\author{ Xudong Qin\inst{2,3}  \ \ \  Eric Madelaine\inst{1,2}
  \ \ \  Min Zhang\inst{3}} 
\institute{Univ. of Nice Sophia Antipolis, CNRS, UMR 7271, 06900 Sophia Antipolis, France
	\and INRIA Sophia Antipolis M\'edit\'erann\'ee, BP 93, 06902 Sophia Antipolis, France
\and Shanghai Key Laboratory of Trustworthy Computing, ECNU, China}
\date{}                                           % Activate to display a given date or no date

@@

\begin{abstract}
  \TODO{not good for a FM, (to be rewriten by Eric)}

  Open pNets are used to model the behavior of open
  synchronous or asynchronous systems expressed in various calculi
  or languages. They are endowed with a symbolic operational semantics
  in terms of so-called ``Open Automata''. We implement an algorithm
  computing this semantics, and building predicates expressing the
  synchronization conditions allowing some combination of events in
  the pNet system. Checking such predicates requires symbolic
  reasoning over first order logics, and specific functions of each
  action algebra. To reduce the complexity of the generated open
  automata, we use the Z3 SMT engine to check satisfiability of the
  predicates, and minimize the number of significant symbolic
  transitions.

@@

%The preceding definition relies on the auxiliary functions below:

%\begin{definition}[Sorts, Holes, Leaves of pNets]
%  \begin{itemize}
%  \item The sort of a pNet is its signature, i.e. the set of actions it can
%perform. In the definition of sorts, we do not need to distinguish
%input variables (that specify the dataflow within LTSs), so for
%computing LTS sorts, we use a substitution operator\footnote{$\subst{y_k\gets x_k}^{k\in K}$ is the parallel substitution 
%operation.} to remove the
%\emph{input marker} of variables. Formally:
%\[
%\begin{array}{l}
%\Sort(\mylangle S,s_0, \to\myrangle) = \{\alpha\subst{x \gets ?x| 
%x\in\symb{iv}(\alpha)}|s \xrightarrow{\langle \alpha,~e_b,~(x_j\!:= {e}_j)^{j\in
%		J}\rangle} s'\in \to \} \\ \Sort(\mylangle
%\set{\pNet}, \set{\symb{S}}, \set{\symb{SV}}\myrangle) =\{\alpha'_k |\, \alpha_j^{j\in J_k}\to\alpha'_k\in\set{\symb{SV}}\}
%\end{array}
%\]
%
%\item
%The set of holes of a pNet is defined inductively; the sets of holes
%in a pNet node and its subnets are all disjoint:
%  \[\begin{array}{l}
%\Holes(\mylangle S,s_0, \to\myrangle) \!=\! \emptyset \\
%\Holes(\mylangle \pNet_i^{i\in I}\!,S_j^{j\in J}\!, \overline{\symb{SV}}\myrangle) =J\cup{\displaystyle \bigcup_{i\in I}\Holes(\pNet_i)}\\
%\forall i\in I.\, \Holes(\pNet_i)\cap J=\emptyset\\
%\forall i_1,i_2\in I.\,i_1\neq i_2\Rightarrow  \Holes(\pNet_{i_1})\cap\Holes(\pNet_{i_2})=\emptyset
%\end{array}\]
%\item
%The set of leaves of a pNet is the set of all pLTSs occurring in the structure, defined inductively as:
%\[\begin{array}{l}
%\Leaves(\mylangle S,s_0, \to\myrangle) \!=\! \{ \mylangle S,s_0, \to\myrangle \}\\
%\Leaves(\mylangle \pNet_i^{i\in I}\!,S_j^{j\in J}\!, \overline{\symb{SV}}\myrangle) = {\displaystyle \bigcup_{i\in I}\Leaves(\pNet_i)}
%\end{array}\]
%\end{itemize}
%\end{definition}
%
%A pNet $Q$ is \emph{closed} if it has no hole: $\Holes(Q)=\emptyset$; else it
%is said to be \emph{open}.

%% \begin{figure}[t]
%% \begin{minipage}{6.1cm}

%%   \includegraphics[width=\linewidth]{XFIG/Enable1}
%%   \\[1.3ex]
%%  \includegraphics[width=\linewidth]{XFIG/Enable2}
%%   \caption{Two pNet encodings for  Enable }  \label{schema:enable-pnets}
%% \end{minipage}
%%   \hspace{2mm}
%% \begin{minipage}{6cm}
%%   \includegraphics[width=\linewidth]{XFIG/P-QR}
%%   \caption{Composed pNet for ``P>>(Q>>R)''}  \label{schema:enable-composed}
%% \end{minipage}

%% \end{figure}

\section{BIP architectures, and their encodings into pNets}
\label{section:BIParchitectures}

\section{Running example}
\label{section:FailureTimerMax}

Several publications \cite{HMZ:PDP15,henrio:Forte2016} already have introduced
many examples of pNets, encoding
operators of various classical process algebras, or more complex
synchronization structures in distributed or parallel languages.

@@

\paragraph{Predicates:}
Let
$\mylangle\overline{\pNet},\overline{S},\symb{SV}_k^{k\in K} \myrangle$
be a pNet. Consider a synchronization vector $SV_k$, for $k\in K$. We build a
predicate $\MkPred$ relating
the actions of the involved sub-pNets and the resulting actions. This predicate verifies:
\[\MkPred(SV_k, a_i^{i\in I}, b_j^{j\in J}, v)\Leftrightarrow
\begin{array}{l}
\exists {(a'_i)}^{i\in I},
{(b'_j)}^{j\in J},v'.\, SV_k={(a'_i)}^{i\in I}, {(b'_j)}^{j\in J}\rightarrow v'
\\~~\land
\forall i\in I.\, a_i=a'_i\land \forall j \in J.\, b_j=b'_j \land v=v'
\end{array}\]
%% In any other case (if the action families do not match or if there is no valuation of
%% variables such that the above formula can be ensured) the predicate is undefined.

@@

%\begin{example}\emph{An open-transition.}
%  \label{OT:lotos-composed}
%
%The \texttt{LOTOS} pNet of Fig. \ref{schema:lotos-pnet} has 2
%controllers and 2 holes. We will show its full open Automaton in
%section \ref{section:full-result}. We give one of its transitions here:
%
% \smallskip
% $ ot_2  = \openrule{
%                       \{0 \xrightarrow{\delta}_{C_1} 0, 0 \xrightarrow{acc(x)}_{C_2} 1\}, ~~
%                       \{\xrightarrow{hb_{12}}_P\}, ~~ 
%                        [s_0=0 \!\land\! hb_{12}=\delta(x)], ~~ 
%                       \{s_0 \xleftarrow{} 1\}     
%                      }
%     {\ostate{00} \xrightarrow{\underline{\delta(x)}} \ostate{01}}$
%\end{example}
%
%The open transition $ot_2$ involves the possible actions from the process $P$ and prefix expression $acc(x);Q$. The hole behavior, represented by variable $hb_{12}$ must be equal to the exit action $\delta(x)$, and the accept event $acc(x)$ will capture the same value $x$.
%
%The transition in the conclusion of this rule starts in
%state $\ostate{00}$, where both $C_1$ and $C_2$ are in their initial
%states. It ends in a different structured state, because in $ot_2$
%both pLTS conduct transitions, showing that the control has gone from
%processes $P$ to $(acc(x);Q$, consuming actions $\delta(x)$ and
%$acc(x)$. Controller $C_1$ has only one state, but its state variable
%$s_0$ has changed to 1.

\paragraph{Structural Semantic Rules:}
Now we build the semantics of an open pNet as an open automaton where
LTSs are the pLTSs at 
the pNet leaves, and the states are structured as in
the previous section. 
%given by Definition~\ref{def-states}.
To build an open transition one first
 projects the global state into states of the leaves, then applies
pLTS transitions on these states, and compose them with actions of
holes using synchronisation vectors. %The pNet
%structure does not appear in the open-automaton, only the
%set of Holes and the set of Leaves.

@@

\subsection{Simplification of Predicates}
\QIN{
We apply several structural rules to generate the predicates restricting the composition of the subnets by synchronization vectors.
The predicates may contain some redundancy if there are variables in middle level are matching other variables both in lower and higher level at the same time. 
}

@@

\section{More details on the generation algorithm}
\subsection{Algebra specification}
\subsection{Term algebra: type system and static semantics}
\subsection{Fresh variables}

>>>>

\title{Using SMT engine to generate Symbolic Automata\thanks{This work was partially 
funded by the Associated Team FM4CPS
  between INRIA and ECNU, Shanghai}}
\author{ Xudong Qin\inst{1,2}  \ \ \  Simon Bliudze\inst{3} \ \ \ Eric Madelaine\inst{1}
  \ \ \  Min Zhang\inst{2}} 
\institute{Universit\'e C\^ote d'Azur, Inria, CNRS, I3S, 06902 Sophia Antipolis, France
        \and Shanghai Key Laboratory of Trustworthy Computing, ECNU, China
\and INRIA Lille }
\date{}                                           % Activate to display a given date or no date

@@

\begin{abstract}
Open pNets are used to model the behavior of open systems, both
synchronous or asynchronous, expressed in various calculi or
languages. They are endowed with a symbolic operational semantics in
terms of so-called ``Open Automata''. This allows to check properties of
such systems in a compositionnal manner. 
We implement an algorithm computing this semantics,
building predicates expressing the synchronization conditions
between the events of the pNet sub-systems. Checking such
predicates requires symbolic reasoning over first order logics, but
also over application-specific data. We use the Z3 SMT engine to check
satisfiability of the predicates, and prune the open automaton of its
unsatisfiable transitions.
As an industrial oriented use-case, we use so-called "architectures"
of BIP systems, that have been used to specify the control software of
ESA microsatellites. We use pNets to encode BIP architectures,
extended with explicite data in synchronisations, and we compute their
open automaton semantics. Cet automate peut alors être utilisé pour
monter les propriétés comme, \TODO{ par exemple ..}

@@

\section{BIP architectures, and their encodings into pNets}
\label{section:BIParchitectures}
\TODO{Simon: can you do this in less than one page ? If you prefer,
  merge with ``Running example'' may be easier to describe the syntax}

\section{Running example}
\label{section:FailureTimerMax}
\TODO{1.5 page including figures ?}

%% Several publications \cite{HMZ:PDP15,henrio:Forte2016} already have introduced
%% many examples of pNets, encoding
%% operators of various classical process algebras, or more complex
%% synchronization structures in distributed or parallel languages.

\TODO{Should have here both the original BIP drawing (Simon), and the
  pNet encoding (fig by Eric)}

@@

\paragraph{Predicates:}
Let
$\mylangle\overline{\pNet},\overline{S},\symb{SV}_k^{k\in K} \myrangle$
be a pNet. Consider a synchronization vector $SV_k$, for $k\in K$. We build a
predicate $\MkPred$ relating
the actions of the involved sub-pNets and the resulting actions. This predicate verifies:
\[\MkPred(SV_k, a_i^{i\in I}, b_j^{j\in J}, v)\Leftrightarrow
\begin{array}{l}
\exists {(a'_i)}^{i\in I},
{(b'_j)}^{j\in J},v'.\, SV_k={(a'_i)}^{i\in I}, {(b'_j)}^{j\in J}\rightarrow v'
\\~~\land
\forall i\in I.\, a_i=a'_i\land \forall j \in J.\, b_j=b'_j \land v=v'
\end{array}\]

@@

\begin{example}\emph{An open-transition.}
  \label{OT:lotos-composed}
%
%The \texttt{LOTOS} pNet of Fig. \ref{schema:lotos-pnet} has 2
%controllers and 2 holes. We will show its full open Automaton in
%section \ref{section:full-result}. We give one of its transitions here:
%
% \smallskip
% $ ot_2  = \openrule{
%                       \{0 \xrightarrow{\delta}_{C_1} 0, 0 \xrightarrow{acc(x)}_{C_2} 1\}, ~~
%                       \{\xrightarrow{hb_{12}}_P\}, ~~ 
%                        [s_0=0 \!\land\! hb_{12}=\delta(x)], ~~ 
%                       \{s_0 \xleftarrow{} 1\}     
%                      }
%     {\ostate{00} \xrightarrow{\underline{\delta(x)}} \ostate{01}}$
\end{example}
%
%The open transition $ot_2$ involves the possible actions from the process $P$ and prefix expression $acc(x);Q$. The hole behavior, represented by variable $hb_{12}$ must be equal to the exit action $\delta(x)$, and the accept event $acc(x)$ will capture the same value $x$.
%
%The transition in the conclusion of this rule starts in
%state $\ostate{00}$, where both $C_1$ and $C_2$ are in their initial
%states. It ends in a different structured state, because in $ot_2$
%both pLTS conduct transitions, showing that the control has gone from
%processes $P$ to $(acc(x);Q$, consuming actions $\delta(x)$ and
%$acc(x)$. Controller $C_1$ has only one state, but its state variable
%$s_0$ has changed to 1.

\paragraph{Structural Semantic Rules:}
\TODO{Do we push the rules in the appendix, and leave only a short high
  level desciption here ?}

Now we build the semantics of an open pNet as an open automaton where
LTSs are the pLTSs at 
the pNet leaves, and the states are structured as in
the previous section. 
%given by Definition~\ref{def-states}.
To build an open transition one first
 projects the global state into states of the leaves, then applies
pLTS transitions on these states, and compose them with actions of
holes using synchronisation vectors. %The pNet
%structure does not appear in the open-automaton, only the
%set of Holes and the set of Leaves.

@@

\subsection{Simplification of Predicates}
\TODO{[Eric]: Motivate, and refs  -- reduce the section here to approx
  half a page} 
\QIN{
We apply several structural rules to generate the predicates restricting the composition of the subnets by synchronization vectors.
The predicates may contain some redundancy if there are variables in middle level are matching other variables both in lower and higher level at the same time. 
}

@@

\section{More details on the generation algorithm}

\subsection{Algebra specification}
\subsection{Term algebra: type system and static semantics}
\subsection{Fresh variables}
